---
title: "House Pricing"
author: "Aslınur Alkan"
date: "2025-07-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this report we perform exploratory and statistical analysis on a random sample of 30 observations from the Ames Housing dataset.  The data are read from **AmesHousing.csv** and pre‑processed before analysis.  The report covers distribution of sale prices, measures of central tendency and variation, normality tests, confidence intervals, hypothesis tests, correlation, regression and other classical statistical methods.

## Code and Analysis

The following R code block reads the Ames Housing dataset, selects relevant features, draws a random sample, and performs various statistical analyses.  Inline comments explain what each section does.  All required packages are loaded at the beginning of the chunk.

```{r analysis, echo=TRUE}
# Load required packages
library(readr)
library(dplyr)
library(e1071)

# Read the dataset
AmesHousing <- read_csv("~/Downloads/AmesHousing.csv")

# Select features
housing_small <- AmesHousing %>%
  select(
    SalePrice,            # sale price
    `Gr Liv Area`,        # living area
    `TotRms AbvGrd`,      # total number of rooms
    `Year Built`,         # year built
    `Overall Qual`,       # overall quality
    `House Style`,        # house type
    Neighborhood,         # neighborhood
    `Garage Cars`,        # garage capacity
    `Full Bath`           # number of bathrooms
  )

# Select 30 random samples
set.seed(42)
housing_sample <- housing_small %>% sample_n(30)
write.csv(housing_sample, "housing_sample_30.csv", row.names = FALSE)

### 3- Frequency Distributions

# Number of classes (Sturges' rule)
n <- nrow(housing_sample)
k <- ceiling(1 + 3.3 * log10(n))
cat("Appropriate number of classes:", k, "\n")

# Price range and class width
range_sp <- range(housing_sample$SalePrice)
range_value <- diff(range_sp)
width <- ceiling(range_value / k)
breaks <- seq(range_sp[1], range_sp[2] + width, by = width)

# Histogram and frequency polygon
hist(housing_sample$SalePrice,
     breaks = breaks,
     main = "Histogram of Sale Prices",
     xlab = "Sale Price",
     col = "blue")

hist_data <- hist(housing_sample$SalePrice, breaks = breaks, plot = FALSE)
midpoints <- hist_data$mids
counts <- hist_data$counts
plot(midpoints, counts,
     type = "o",
     col = "red",
     main = "Frequency Polygon of Sale Prices",
     xlab = "Sale Price",
     ylab = "Frequency")

### 4- Data Description

# Measures of central tendency
mean_price <- mean(housing_sample$SalePrice)
median_price <- median(housing_sample$SalePrice)
estimated_mode <- midpoints[which.max(counts)]
cat("Mean:", mean_price, "\n")
cat("Median:", median_price, "\n")
cat("Estimated Mode from histogram:", estimated_mode, "\n")

# Measures of variation
sd_price <- sd(housing_sample$SalePrice)
cv_price <- sd_price / mean_price
cat("Standard deviation:", sd_price, "\n")
cat("Coefficient of Variation:", cv_price, "\n")

# Measures of position
z_scores <- scale(housing_sample$SalePrice)
quantile(housing_sample$SalePrice, probs = seq(0, 1, 0.1))
summary(housing_sample$SalePrice)
quartiles <- quantile(housing_sample$SalePrice, probs = c(0.25, 0.5, 0.75))
boxplot(housing_sample$SalePrice,
        main = "Boxplot of Sale Prices",
        ylab = "Sale Price",
        col = "pink")

# Outliers
box_stats <- boxplot.stats(housing_sample$SalePrice)
outliers <- box_stats$out
cat("Outliers:", outliers, "\n")

### 5- Normal Distribution

hist(housing_sample$SalePrice,
     breaks = breaks,
     probability = TRUE,
     col = "lightblue",
     main = "Histogram of Sale Prices with Normal Curve",
     xlab = "Sale Price")
curve(dnorm(x, mean = mean_price, sd = sd_price),
      col = "darkblue",
      lwd = 2,
      add = TRUE)

shapiro_test <- shapiro.test(housing_sample$SalePrice)
shapiro_test
skewness(housing_sample$SalePrice)

# Central Limit Theorem example
set.seed(42)
sample_means <- replicate(1000, {
  sample <- sample(housing_small$SalePrice, size = 30, replace = TRUE)
  mean(sample)
})
hist(sample_means,
     probability = TRUE,
     main = "Sample Means Histogram (Central Limit Theorem)",
     xlab = "Sample Means",
     col = "blue")
curve(dnorm(x, mean = mean(sample_means), sd = sd(sample_means)),
      add = TRUE,
      col = "red",
      lwd = 2)

### 6- Confidence Intervals & Sampling Size

df <- n - 1
t_calculated <- qt(0.975, df)
error_margin <- qt(0.975, df) * sd_price / sqrt(n)
lower_bound <- mean_price - error_margin
upper_bound <- mean_price + error_margin

p_hat <- mean(housing_sample$`Garage Cars` >= 1)
z <- qnorm(0.975)
error_prop <- z * sqrt((p_hat * (1 - p_hat)) / n)
ci_prop <- c(p_hat - error_prop, p_hat + error_prop)

cat("95% confidence interval for mean sale price:\n")
cat("[", lower_bound, ",", upper_bound, "]\n")
cat("95% Confidence Interval for Proportion of Houses with Garage:", ci_prop , "\n")

### 7- Hypothesis Testing

p_value <- 2 * pt(-abs(t_calculated), df)
cat("p-value:", p_value, "\n")
mu_0 <- 200000
t_critical <- qt(1 - 0.05 / 2, df)
cat("Test Statistic (t_calculated):", t_calculated, "\n")
cat("Critical Value (±t):", -t_critical, "to", t_critical, "\n")

### 8- Testing the Difference Between Two Means

subset_data <- housing_sample %>%
  filter(`House Style` %in% c("1Story", "2Story"))
t.test(SalePrice ~ `House Style`, data = subset_data)
garage_table <- table(subset_data$`House Style`, subset_data$`Garage Cars` >= 1)
x1 <- garage_table["1Story", "TRUE"]
x2 <- garage_table["2Story", "TRUE"]
n1 <- sum(garage_table["1Story", ])
n2 <- sum(garage_table["2Story", ])
prop.test(x = c(x1, x2), n = c(n1, n2))
var.test(SalePrice ~ `House Style`, data = subset_data)

### 9- Correlation and Regression

correlation <- cor(housing_sample$`Gr Liv Area`, housing_sample$SalePrice)
correlation
cor.test(housing_sample$`Gr Liv Area`, housing_sample$SalePrice)
model <- lm(SalePrice ~ `Gr Liv Area`, data = housing_sample)
summary(model)
plot(housing_sample$`Gr Liv Area`, housing_sample$SalePrice,
     main = "Scatter Plot with Regression Line",
     xlab = "Gr Liv Area", ylab = "Sale Price",
     col = "blue", pch = 19)
abline(model, col = "red", lwd = 2)
plot(model$residuals,
     main = "Residual Plot",
     ylab = "Residuals",
     xlab = "Observation Index",
     col = "purple", pch = 16)
abline(h = 0, col = "red", lty = 2)
r_squared <- summary(model)$r.squared
adjusted_r_squared <- summary(model)$adj.r.squared
standard_error <- summary(model)$sigma
cat("R-squared:", r_squared, "\n")
cat("Adjusted R-squared:", adjusted_r_squared, "\n")
cat("Standard Error of Estimate:", standard_error, "\n")

### 10- Chi-Square Tests

garage_counts <- table(housing_sample$`Garage Cars`)
chisq.test(garage_counts)
garage_binary <- ifelse(housing_sample$`Garage Cars` >= 1, "Yes", "No")
contingency_table <- table(housing_sample$Neighborhood, garage_binary)
chisq.test(contingency_table)

### 11- Analysis of Variance (ANOVA)

housing_sample$HouseStyle <- as.factor(housing_sample$`House Style`)
oneway <- aov(SalePrice ~ HouseStyle, data = housing_sample)
summary(oneway)
TukeyHSD(oneway)

### 12- Non‑parametric Tests

spearman <- cor.test(housing_sample$`Gr Liv Area`, housing_sample$SalePrice, method = "spearman")
spearman
wilcox <- wilcox.test(housing_sample$`Gr Liv Area`, housing_sample$`TotRms AbvGrd`, paired = TRUE)
wilcox
```

## Conclusion

The analysis above explores the distribution of sale prices in the Ames Housing sample, calculates measures of central tendency and variation, assesses normality, constructs confidence intervals, performs hypothesis testing, examines the relationship between living area and sale price via simple linear regression, and applies chi‑square, ANOVA and non‑parametric tests where appropriate.  These results provide a comprehensive statistical overview of the selected housing data.
